---
title: Columbia River (Snake) Summer Steelhead SAR Models & Forecasts
author: Thomas Buehrens (tbuehrens@dfw.wa.gov)
output:
  html_document:
    code_folding: hide
    fig_caption: yes
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

<script>
   $(document).ready(function() {
     $head = $('#header');
     $head.prepend('<img src=\"https://privatelands.wdfw.wa.gov/wdfwlogo_clrnotxt.png"\" style=\"float: right;width: 150px;\"/>')
   });
</script>

***

Last Updated `r format(Sys.time(), '%m/%d/%Y')`.

***


# Setup
All analyses require R software [**(link)**](https://cran.r-project.org/) for data retrieval, data processing, and summarizing model results. Here we configure R to perform our analysis and generate our outputs
```{r set_options, echo = TRUE, message = FALSE}
options(width = 100)
knitr::opts_chunk$set(message = FALSE)
set.seed(123)
```

We also need a couple of helper functions which we will load from the functions folder, which we will load using the walk() function from the purrr package (which we will install if it is not already installed).
```{r load_funcs, message = FALSE, warning = FALSE,results = "hide"}
wd_functions<-"functions"
sapply(FUN = source, paste(wd_functions, list.files(wd_functions), sep="/"))

logit<-function(x){log(x/(1-x))}
ilogit<-function(x){exp(x)/(1+exp(x))}
```

Here we will load & install packages we need to use (needs internet connection if packages not already installed)
```{r load_packages, message = FALSE, warning = FALSE,results = "hide"}
packages_list<-c("tidyverse"
                 ,"forecast"
                 ,"mgcv"
                 ,"ggplot2"
                 ,"kableExtra"
                 ,"brms"
                 ,"modelr"
                 ,"gamm4"
                 ,"gridExtra"
                 ,"rvest"
                 ,"janitor"
                 )
install_or_load_pack(pack = packages_list)
```

# Load Data
```{r load_data, message = FALSE, warning = FALSE,results = "show"}
NPGO<-read_table("http://www.o3d.org/npgo/npgo.php",skip=29,col_names=F,comment="#")%>%
  filter(!is.na(X2))%>%
  dplyr::rename(year=X1,Month=X2,NPGO=X3)%>%
  mutate(year=as.numeric(year))%>%
  group_by(year)%>%
  add_tally()%>%
  group_by(year)%>%
  dplyr::summarise(NPGO=mean(NPGO))

SAR_data<-read_html("https://www.cbr.washington.edu/dart/wrapper?type=php&fname=pitsaresu_1669129113_863.csv")%>%
  html_nodes("body")%>%
  html_text()%>%
  read_lines(skip_empty_rows = T)%>%
  tibble()%>%
  setNames("lines")%>%
  filter(grepl("year",lines) | grepl("20",lines) & !grepl("Generated",lines) & !grepl("DART Data Citation",lines))%>%
  as_vector()%>%
  str_split(",")%>%
  as.data.frame()%>%
  as.matrix()%>%
  t()%>%
  as_tibble()%>%
  row_to_names(row_number = 1)%>%
  type.convert()

 
dat<-SAR_data%>%
  dplyr::select(year,juvCount,ocean1Count,ocean2Count)%>%
  full_join(NPGO)%>%
  arrange(year)%>%
  mutate(even_odd=year%%2,
         juvCount = ifelse(!is.na(juvCount),juvCount,0),
         logit_p1 = logit(ocean1Count/juvCount),
         logit_p2 = ifelse(ocean2Count>1,logit(ocean2Count/juvCount),NA),
         oceanALLCount = ocean1Count + ocean2Count,
         logit_pALL = ifelse(ocean2Count>1,logit(oceanALLCount/juvCount),NA),
         )
```

# Ocean age 1 Forecast
```{r OA1_forecasts, message = FALSE, warning = FALSE, results = "hide"}
m1<-gam(cbind(ocean1Count,juvCount-ocean1Count) ~ s(year,bs="bs",m=1) + even_odd + NPGO,data=dat%>%filter(juvCount>=1),family="binomial")
summary(m1)
plot(m1)
AIC(m1)

m2<-gamm(cbind(ocean1Count,juvCount-ocean1Count) ~  s(year,bs="bs",m=1) + even_odd + NPGO, correlation=corARMA(q=1),data=dat%>%filter(juvCount>=1),family="binomial")
summary(m2$lme)
pred2<-tibble(pred2=predict(m2$gam,newdata=dat%>%filter(juvCount>=1),type = "response"))
AIC(m2$lme)
plot(m2$gam)


m4<-brm(ocean1Count|trials(juvCount) ~  gp(year) + even_odd + NPGO 
        ,data=dat%>%filter(juvCount>=1)
        ,family="binomial"
        #,prior=set_prior(horseshoe(1, scale_global = 0.5))
        ,prior = set_prior(horseshoe(df = 1, par_ratio = 0.8), class="b")
        ,chains = 4
        ,cores = 4
        ,iter = 2000
        ,warmup = 1000
        ,control=list(adapt_delta = 0.995)
)

m5<-brm(ocean1Count|trials(juvCount) ~  s(year,bs="bs",m=1) + arma(p=1,q=1,cov=T) + even_odd + NPGO 
        ,data=dat%>%filter(juvCount>=1)
        ,family="binomial"
        #,prior=set_prior(horseshoe(1, scale_global = 0.5))
        ,prior = set_prior(horseshoe(df = 1, par_ratio = 0.8), class="b")
        ,chains = 4
        ,cores = 4
        ,iter = 2000
        ,warmup = 1000
        ,control=list(adapt_delta = 0.995)
)

m6<-brm(ocean1Count|trials(juvCount) ~  arma(p=1,q=1,cov=T) + even_odd + NPGO 
        ,data=dat%>%filter(juvCount>=1)
        ,family="binomial"
        #,prior=set_prior(horseshoe(1, scale_global = 0.5))
        ,prior = set_prior(horseshoe(df = 1, par_ratio = 0.8), class="b")
        ,chains = 4
        ,cores = 4
        ,iter = 2000
        ,warmup = 1000
        ,control=list(adapt_delta = 0.995)
)

pred4=posterior_linpred(m4,newdata=dat%>%
                           filter(juvCount>=1))%>%
  ilogit()%>%
  t()%>%
  tibble()%>%
  rowwise()%>%
  summarize(across(everything(),~median(.)))%>%
  setNames("pred4")

pred5=posterior_linpred(m5,newdata=dat%>%
                           filter(juvCount>=1))%>%
  ilogit()%>%
  t()%>%
  tibble()%>%
  rowwise()%>%
  summarize(across(everything(),~median(.)))%>%
  setNames("pred5")

pred6=posterior_linpred(m6,newdata=dat%>%
                           filter(juvCount>=1))%>%
  ilogit()%>%
  t()%>%
  tibble()%>%
  rowwise()%>%
  summarize(across(everything(),~median(.)))%>%
  setNames("pred6")


newdat<-dat%>%
  filter(juvCount>=1)%>%
  add_predictions(m1,type="response")%>%
  bind_cols(pred2)%>%
  bind_cols(pred4=pred4)%>%
  bind_cols(pred5=pred5)%>%
  bind_cols(pred6=pred6)
  

p1<-ggplot(newdat,aes(x=year,y=pred))+
  geom_line()+
  geom_line(mapping=aes(x=year,y=pred2),color="red")+
  geom_line(mapping=aes(x=year,y=pred4),color="orange")+
  geom_line(mapping=aes(x=year,y=pred5),color="green")+
  geom_line(mapping=aes(x=year,y=pred6),color="purple")+
  geom_point(mapping=aes(x=year,y=ilogit(logit_p1)))+
  ylim(0,NA)+
  ylab("Ocean Age 1 Survival")
```

# Ocean age 2 Forecast
```{r OA2_forecasts, message = FALSE, warning = FALSE, results = "hide"}
m1<-gam(cbind(ocean2Count,juvCount-ocean2Count) ~ s(year,bs="bs",m=1) + even_odd + NPGO,data=dat%>%filter(juvCount>=1),family="binomial")
summary(m1)
plot(m1)
AIC(m1)

m2<-gamm(cbind(ocean2Count,juvCount-ocean2Count) ~  s(year,bs="bs",m=1) + even_odd + NPGO, correlation=corARMA(q=1),data=dat%>%filter(juvCount>=1),family="binomial")
summary(m2$lme)
pred2<-tibble(pred2=predict(m2$gam,newdata=dat%>%filter(juvCount>=1),type = "response"))
AIC(m2$lme)
plot(m2$gam)


m4<-brm(ocean2Count|trials(juvCount) ~  gp(year) + even_odd + NPGO 
        ,data=dat%>%filter(juvCount>=1)
        ,family="binomial"
        #,prior=set_prior(horseshoe(1, scale_global = 0.5))
        ,prior = set_prior(horseshoe(df = 1, par_ratio = 0.8), class="b")
        ,chains = 4
        ,cores = 4
        ,iter = 2000
        ,warmup = 1000
        ,control=list(adapt_delta = 0.995)
)

m5<-brm(ocean2Count|trials(juvCount) ~  s(year,bs="bs",m=1) + arma(p=1,q=1,cov=T) + even_odd + NPGO 
        ,data=dat%>%filter(juvCount>=1)
        ,family="binomial"
        #,prior=set_prior(horseshoe(1, scale_global = 0.5))
        ,prior = set_prior(horseshoe(df = 1, par_ratio = 0.8), class="b")
        ,chains = 4
        ,cores = 4
        ,iter = 2000
        ,warmup = 1000
        ,control=list(adapt_delta = 0.995)
)

m6<-brm(ocean2Count|trials(juvCount) ~  arma(p=1,q=1,cov=T) + even_odd + NPGO 
        ,data=dat%>%filter(juvCount>=1)
        ,family="binomial"
        #,prior=set_prior(horseshoe(1, scale_global = 0.5))
        ,prior = set_prior(horseshoe(df = 1, par_ratio = 0.8), class="b")
        ,chains = 4
        ,cores = 4
        ,iter = 2000
        ,warmup = 1000
        ,control=list(adapt_delta = 0.995)
)

pred4=posterior_linpred(m4,newdata=dat%>%
                           filter(juvCount>=1))%>%
  ilogit()%>%
  t()%>%
  tibble()%>%
  rowwise()%>%
  summarize(across(everything(),~median(.)))%>%
  setNames("pred4")

pred5=posterior_linpred(m5,newdata=dat%>%
                           filter(juvCount>=1))%>%
  ilogit()%>%
  t()%>%
  tibble()%>%
  rowwise()%>%
  summarize(across(everything(),~median(.)))%>%
  setNames("pred5")

pred6=posterior_linpred(m6,newdata=dat%>%
                           filter(juvCount>=1))%>%
  ilogit()%>%
  t()%>%
  tibble()%>%
  rowwise()%>%
  summarize(across(everything(),~median(.)))%>%
  setNames("pred6")


newdat2<-dat%>%
  filter(juvCount>=1)%>%
  add_predictions(m1,type="response")%>%
  bind_cols(pred2)%>%
  bind_cols(pred4=pred4)%>%
  bind_cols(pred5=pred5)%>%
  bind_cols(pred6=pred6)
  

p2<-ggplot(newdat2,aes(x=year,y=pred))+
  geom_line()+
  geom_line(mapping=aes(x=year,y=pred2),color="red")+
  geom_line(mapping=aes(x=year,y=pred4),color="orange")+
  geom_line(mapping=aes(x=year,y=pred5),color="green")+
  geom_line(mapping=aes(x=year,y=pred6),color="purple")+
  geom_point(mapping=aes(x=year,y=ilogit(logit_p2)))+
  ylim(0,NA)+
  ylab("Ocean Age 2 Survival")
```

# All Ocean age  Forecast
```{r All_OA_forecasts, message = FALSE, warning = FALSE, results = "hide"}
m1<-gam(cbind(oceanALLCount,juvCount-oceanALLCount) ~ s(year,bs="bs",m=1) + even_odd + NPGO,data=dat%>%filter(juvCount>=1),family="binomial")
summary(m1)
plot(m1)
AIC(m1)

m2<-gamm(cbind(oceanALLCount,juvCount-oceanALLCount) ~  s(year,bs="bs",m=1) + even_odd + NPGO, correlation=corARMA(q=1),data=dat%>%filter(juvCount>=1),family="binomial")
summary(m2$lme)
pred2<-tibble(pred2=predict(m2$gam,newdata=dat%>%filter(juvCount>=1),type = "response"))
AIC(m2$lme)
plot(m2$gam)


m4<-brm(oceanALLCount|trials(juvCount) ~  gp(year) + even_odd + NPGO 
        ,data=dat%>%filter(juvCount>=1)
        ,family="binomial"
        #,prior=set_prior(horseshoe(1, scale_global = 0.5))
        ,prior = set_prior(horseshoe(df = 1, par_ratio = 0.8), class="b")
        ,chains = 4
        ,cores = 4
        ,iter = 2000
        ,warmup = 1000
        ,control=list(adapt_delta = 0.995)
)

m5<-brm(oceanALLCount|trials(juvCount) ~  s(year,bs="bs",m=1) + arma(p=1,q=1,cov=T) + even_odd + NPGO 
        ,data=dat%>%filter(juvCount>=1)
        ,family="binomial"
        #,prior=set_prior(horseshoe(1, scale_global = 0.5))
        ,prior = set_prior(horseshoe(df = 1, par_ratio = 0.8), class="b")
        ,chains = 4
        ,cores = 4
        ,iter = 2000
        ,warmup = 1000
        ,control=list(adapt_delta = 0.995)
)

m6<-brm(oceanALLCount|trials(juvCount) ~  arma(p=1,q=1,cov=T) + even_odd + NPGO 
        ,data=dat%>%filter(juvCount>=1)
        ,family="binomial"
        #,prior=set_prior(horseshoe(1, scale_global = 0.5))
        ,prior = set_prior(horseshoe(df = 1, par_ratio = 0.8), class="b")
        ,chains = 4
        ,cores = 4
        ,iter = 2000
        ,warmup = 1000
        ,control=list(adapt_delta = 0.995)
)

pred4=posterior_linpred(m4,newdata=dat%>%
                           filter(juvCount>=1))%>%
  ilogit()%>%
  t()%>%
  tibble()%>%
  rowwise()%>%
  summarize(across(everything(),~median(.)))%>%
  setNames("pred4")

pred5=posterior_linpred(m5,newdata=dat%>%
                           filter(juvCount>=1))%>%
  ilogit()%>%
  t()%>%
  tibble()%>%
  rowwise()%>%
  summarize(across(everything(),~median(.)))%>%
  setNames("pred5")

pred6=posterior_linpred(m6,newdata=dat%>%
                           filter(juvCount>=1))%>%
  ilogit()%>%
  t()%>%
  tibble()%>%
  rowwise()%>%
  summarize(across(everything(),~median(.)))%>%
  setNames("pred6")


newdat3<-dat%>%
  filter(juvCount>=1)%>%
  add_predictions(m1,type="response")%>%
  bind_cols(pred2)%>%
  bind_cols(pred4=pred4)%>%
  bind_cols(pred5=pred5)%>%
  bind_cols(pred6=pred6)
  

p3<-ggplot(newdat3,aes(x=year,y=pred))+
  geom_line()+
  geom_line(mapping=aes(x=year,y=pred2),color="red")+
  geom_line(mapping=aes(x=year,y=pred4),color="orange")+
  geom_line(mapping=aes(x=year,y=pred5),color="green")+
  geom_line(mapping=aes(x=year,y=pred6),color="purple")+
  geom_point(mapping=aes(x=year,y=ilogit(logit_pALL)))+
  ylim(0,NA)+
  ylab("ALL Ocean Age Survival")
```

# Results
```{r Results, message = FALSE, warning = FALSE, results = "show"}
grid.arrange(p1,p2,p3,ncol=1)

newdat%>%
  mutate(OA1_SAR=ilogit(logit_p1))%>%
  dplyr::select(year,
                OA1_SAR,
                pred,
                pred2,
                pred4,
                pred5,
                pred6
                )%>%
  mutate(across(-year,~scales::percent(.,accuracy = 0.01)))%>%
  kbl(caption = "Table 1. Ocean Age 1 Survival Observed & Predicted.",digits =1)%>%
  kable_classic(full_width = F, html_font = "Cambria")

newdat2%>%
  mutate(OA2_SAR=ilogit(logit_p2))%>%
  dplyr::select(year,
                OA2_SAR,
                pred,
                pred2,
                pred4,
                pred5,
                pred6
                )%>%
  mutate(across(-year,~scales::percent(.,accuracy = 0.01)))%>%
  kbl(caption = "Table 2. Ocean Age 2 Survival Observed & Predicted.",digits =1)%>%
  kable_classic(full_width = F, html_font = "Cambria")

newdat3%>%
  mutate(ALL_OA_SAR=ilogit(logit_pALL))%>%
  dplyr::select(year,
                ALL_OA_SAR,
                pred,
                pred2,
                pred4,
                pred5,
                pred6
                )%>%
  mutate(across(-year,~scales::percent(.,accuracy = 0.01)))%>%
  kbl(caption = "Table 3. All Ocean Age Survival Observed & Predicted.",digits =1)%>%
  kable_classic(full_width = F, html_font = "Cambria")
  
```