---
title: Columbia River (Snake) Summer Steelhead SAR Models & Forecasts
author: Thomas Buehrens (tbuehrens@dfw.wa.gov)
output:
  html_document:
    code_folding: hide
    fig_caption: yes
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

<script>
   $(document).ready(function() {
     $head = $('#header');
     $head.prepend('<img src=\"https://privatelands.wdfw.wa.gov/wdfwlogo_clrnotxt.png"\" style=\"float: right;width: 150px;\"/>')
   });
</script>

***

Last Updated `r format(Sys.time(), '%m/%d/%Y')`.

***


# Setup
All analyses require R software [**(link)**](https://cran.r-project.org/) (v3.4.3) for data retrieval, data processing, and summarizing model results. Here we configure R to perform our analysis and generate our outputs
```{r set_options, echo = TRUE, message = FALSE}
options(width = 100)
knitr::opts_chunk$set(message = FALSE)
set.seed(123)
```

We also need a couple of helper functions which we will load from the functions folder, which we will load using the walk() function from the purrr package (which we will install if it is not already installed).
```{r load_funcs, message = FALSE, warning = FALSE,results = "hide"}
wd_functions<-"functions"
sapply(FUN = source, paste(wd_functions, list.files(wd_functions), sep="/"))

logit<-function(x){log(x/(1-x))}
ilogit<-function(x){exp(x)/(1+exp(x))}
```

Here we will load & install packages we need to use (needs internet connection if packages not already installed)
```{r load_packages, message = FALSE, warning = FALSE,results = "hide"}
packages_list<-c("tidyverse"
                 ,"forecast"
                 ,"mgcv"
                 ,"ggplot2"
                 ,"kableExtra"
                 ,"brms"
                 ,"modelr"
                 ,"gamm4"
                 ,"gridExtra"
                 ,"rvest"
                 ,"janitor"
                 )
install_or_load_pack(pack = packages_list)
```

# Load Data
```{r load_data, message = FALSE, warning = FALSE,results = "show"}
NPGO<-read_table("http://www.o3d.org/npgo/npgo.php",skip=29,col_names=F,comment="#")%>%
  filter(!is.na(X2))%>%
  dplyr::rename(year=X1,Month=X2,NPGO=X3)%>%
  mutate(year=as.numeric(year))%>%
  group_by(year)%>%
  add_tally()%>%
  group_by(year)%>%
  dplyr::summarise(NPGO=mean(NPGO))

SAR_data<-read_html("https://www.cbr.washington.edu/dart/wrapper?type=php&fname=pitsaresu_1669129113_863.csv")%>%
  html_nodes("body")%>%
  html_text()%>%
  read_lines(skip_empty_rows = T)%>%
  tibble()%>%
  setNames("lines")%>%
  filter(grepl("year",lines) | grepl("20",lines) & !grepl("Generated",lines) & !grepl("DART Data Citation",lines))%>%
  as_vector()%>%
  str_split(",")%>%
  as.data.frame()%>%
  as.matrix()%>%
  t()%>%
  as_tibble()%>%
  row_to_names(row_number = 1)%>%
  type.convert()

 
dat<-SAR_data%>%
  dplyr::select(year,juvCount,ocean1Count,ocean2Count)%>%
  full_join(NPGO)%>%
  arrange(year)%>%
  mutate(even_odd=year%%2,
         juvCount = ifelse(!is.na(juvCount),juvCount,0),
         lag1_NPGO = lag(NPGO),
         lag2_NPGO = lag(NPGO,2),
         diff_NPGO_L2L1 = c(NA,diff(lag1_NPGO-lag2_NPGO)),
         logit_p1 = logit(ocean1Count/juvCount),
         logit_p2 = ifelse(ocean2Count>1,logit(ocean2Count/juvCount),NA),
         oceanALLCount = ocean1Count + ocean2Count,
         logit_pALL = ifelse(ocean2Count>1,logit(oceanALLCount/juvCount),NA),
         )
```

# Ocean age 1 Forecast
```{r OA1_forecasts, message = FALSE, warning = FALSE, results = "asis", include=TRUE}
m1<-gam(cbind(ocean1Count,juvCount-ocean1Count) ~ s(year,bs="bs",m=1) + even_odd + lag1_NPGO,data=dat%>%filter(juvCount>=1),family="binomial")
summary(m1)
plot(m1)
AIC(m1)

m2<-gamm(cbind(ocean1Count,juvCount-ocean1Count) ~  s(year,bs="bs",m=1) + even_odd + lag1_NPGO, correlation=corARMA(q=1),data=dat%>%filter(juvCount>=1),family="binomial")
summary(m2$lme)
preds2<-tibble(preds2=predict(m2$gam,newdata=dat%>%filter(juvCount>=1),type = "response"))
AIC(m2$lme)
plot(m2$gam)

m3<-auto.arima(y=dat%>%filter(juvCount>=1)%>%dplyr::select(logit_p1)%>%pull(),
               xreg=dat%>%filter(juvCount>=1)%>%dplyr::select(even_odd,lag1_NPGO)%>%as.matrix()
)
f<-forecast(m3,h=1,xreg=dat%>%filter(juvCount>=1)%>%dplyr::select(even_odd,lag1_NPGO)%>%as.matrix())
f<-ilogit(c(f$fitted[!is.na(f$fitted)],f$mean[is.na(f$fitted)]))

newdat<-dat%>%
  filter(juvCount>=1)%>%
  add_predictions(m1,type="response")%>%
  bind_cols(preds2)%>%
  bind_cols(tibble(preds3=f))
  

p1<-ggplot(newdat,aes(x=year,y=pred))+
  geom_line()+
  geom_line(mapping=aes(x=year,y=preds2),color="red")+
  geom_line(mapping=aes(x=year,y=preds3),color="blue")+
  geom_point(mapping=aes(x=year,y=ilogit(logit_p1)))+
  ylim(0,NA)


m4<-brm(ocean1Count|trials(juvCount) ~  gp(year) + even_odd + lag1_NPGO #s(year,bs="bs",m=1)
        ,data=dat%>%filter(juvCount>=1)
        ,family="binomial"
        #,prior=set_prior(horseshoe(1, scale_global = 0.5))
        ,prior = set_prior(horseshoe(df = 1, par_ratio = 0.8), class="b")
        ,chains = 4
        ,cores = 4
        ,iter = 2000
        ,warmup = 1000
        ,control=list(adapt_delta = 0.995)
)

preds4=predict(m4,newdata=dat%>%filter(juvCount>=1))%>%
  as_vector()%>%
  as_tibble()%>%
  dplyr::select(Estimate,Q2.5,Q97.5)%>%
  bind_cols(dat%>%
              filter(juvCount>=1)%>%
              dplyr::select(juvCount,year,logit_p1)
  )%>%
  mutate(across(c(Estimate,Q2.5,Q97.5),~./juvCount+1e-09,byrow=T))

ggplot(preds4,aes(x=year,y=Estimate))+
  geom_line()+
  geom_ribbon(aes(ymin=`Q2.5`,ymax=`Q97.5`),fill="cadetblue",alpha=0.5,color=NA)+
  geom_point(mapping=aes(x=year,y=ilogit(logit_p1)))+
  ylim(0,NA)
```

# Ocean age 2 Forecast
```{r OA2_forecasts, message = FALSE, warning = FALSE, results = "asis", include=TRUE}
m1<-gam(cbind(ocean2Count,juvCount-ocean2Count) ~ s(year,bs="bs",m=1) + even_odd +  lag1_NPGO,data=dat%>%filter(juvCount>=1),family="binomial")
summary(m1)
plot(m1)
AIC(m1)

m2<-gamm(cbind(ocean2Count,juvCount-ocean2Count) ~  s(year,bs="bs",m=1) + even_odd +  lag1_NPGO, correlation=corARMA(q=1),data=dat%>%filter(juvCount>=1),family="binomial")
summary(m2$lme)
preds2<-tibble(preds2=predict(m2$gam,newdata=dat%>%filter(juvCount>=1),type = "response"))
AIC(m2$lme)
plot(m2$gam)


m3<-auto.arima(y=dat%>%filter(juvCount>=1)%>%dplyr::select(logit_p2)%>%pull(),
               xreg=dat%>%filter(juvCount>=1)%>%dplyr::select(even_odd, lag1_NPGO)%>%as.matrix()
)
f<-forecast(m3,h=1,xreg=dat%>%filter(juvCount>=1)%>%dplyr::select(even_odd,lag1_NPGO)%>%as.matrix())
f<-ilogit(c(f$fitted[!is.na(f$fitted)],f$mean[is.na(f$fitted)]))

newdat<-dat%>%
  filter(juvCount>=1)%>%
  add_predictions(m1,type="response")%>%
  bind_cols(preds2)%>%
  bind_cols(tibble(preds3=f))


p2<-ggplot(newdat,aes(x=year,y=pred))+
  geom_line()+
  geom_line(mapping=aes(x=year,y=preds2),color="red")+
  geom_line(mapping=aes(x=year,y=preds3),color="blue")+
  geom_point(mapping=aes(x=year,y=ilogit(logit_p2)))+
  ylim(0,NA)+
  xlim(NA,2022)


m4<-brm(ocean2Count|trials(juvCount) ~  gp(year) + even_odd + lag1_NPGO + lag2_NPGO + diff_NPGO_L2L1 #s(year,bs="bs",m=1)
        ,data=dat%>%filter(juvCount>=1)
        ,family="binomial"
        #,prior=set_prior(horseshoe(1, scale_global = 0.5))
        ,prior = set_prior(horseshoe(df = 1, par_ratio = 0.8), class="b")
        ,chains = 4
        ,cores = 4
        ,iter = 2000
        ,warmup = 1000
        ,control=list(adapt_delta = 0.995)
        )

preds4=predict(m4,newdata=dat%>%filter(juvCount>=1))%>%
  as_vector()%>%
  as_tibble()%>%
  dplyr::select(Estimate,Q2.5,Q97.5)%>%
  bind_cols(dat%>%
              filter(juvCount>=1)%>%
              dplyr::select(juvCount,year,logit_p2)
            )%>%
  mutate(across(c(Estimate,Q2.5,Q97.5),~./juvCount+1e-09,byrow=T))

ggplot(preds4,aes(x=year,y=Estimate))+
  geom_line()+
  geom_ribbon(aes(ymin=`Q2.5`,ymax=`Q97.5`),fill="cadetblue",alpha=0.5,color=NA)+
  geom_point(mapping=aes(x=year,y=ilogit(logit_p2)))+
  ylim(0,NA)
```

# All Ocean age  Forecast
```{r All_OA_forecasts, message = FALSE, warning = FALSE, results = "asis", include=TRUE}
m1<-gam(cbind(oceanALLCount,juvCount-oceanALLCount) ~ s(year,bs="bs",m=1) + even_odd + lag1_NPGO + lag2_NPGO + diff_NPGO_L2L1,data=dat%>%filter(juvCount>=1),family="binomial")
summary(m1)
plot(m1)
AIC(m1)

m2<-gamm(cbind(oceanALLCount,juvCount-oceanALLCount) ~  s(year,bs="bs",m=1) + even_odd +  lag1_NPGO + lag2_NPGO + diff_NPGO_L2L1, correlation=corARMA(q=1),data=dat%>%filter(juvCount>=1),family="binomial")
summary(m2$lme)
preds2<-tibble(preds2=predict(m2$gam,newdata=dat%>%filter(juvCount>=1),type = "response"))
AIC(m2$lme)
plot(m2$gam)


m3<-auto.arima(y=dat%>%filter(juvCount>=1)%>%dplyr::select(logit_pALL)%>%pull(),
               xreg=dat%>%filter(juvCount>=1)%>%dplyr::select(even_odd, lag1_NPGO, lag2_NPGO, diff_NPGO_L2L1)%>%as.matrix()
)
f<-forecast(m3,h=1,xreg=dat%>%filter(juvCount>=1)%>%dplyr::select(even_odd,lag1_NPGO, lag2_NPGO, diff_NPGO_L2L1)%>%as.matrix())
f<-ilogit(c(f$fitted[!is.na(f$fitted)],f$mean[is.na(f$fitted)]))

newdat<-dat%>%
  filter(juvCount>=1)%>%
  add_predictions(m1,type="response")%>%
  bind_cols(preds2)%>%
  bind_cols(tibble(preds3=f))



p3<-ggplot(newdat,aes(x=year,y=pred))+
  geom_line()+
  geom_line(mapping=aes(x=year,y=preds2),color="red")+
  geom_line(mapping=aes(x=year,y=preds3),color="blue")+
  geom_point(mapping=aes(x=year,y=ilogit(logit_pALL)))+
  ylim(0,NA)+
  xlim(NA,2022)

print(p3)

m4<-brm(oceanALLCount|trials(juvCount) ~  gp(year) + even_odd + lag1_NPGO + lag2_NPGO + diff_NPGO_L2L1 #s(year,bs="bs",m=1)
        ,data=dat%>%filter(juvCount>=1)
        ,family="binomial"
        #,prior=set_prior(horseshoe(1, scale_global = 0.5))
        ,prior = set_prior(horseshoe(df = 1, par_ratio = 0.8), class="b")
        ,chains = 4
        ,cores = 4
        ,iter = 2000
        ,warmup = 1000
        ,control=list(adapt_delta = 0.995)
)

preds4=predict(m4,newdata=dat%>%filter(juvCount>=1))%>%
  as_vector()%>%
  as_tibble()%>%
  dplyr::select(Estimate,Q2.5,Q97.5)%>%
  bind_cols(dat%>%
              filter(juvCount>=1)%>%
              dplyr::select(juvCount,year,logit_pALL)
  )%>%
  mutate(across(c(Estimate,Q2.5,Q97.5),~./juvCount+1e-09,byrow=T))

ggplot(preds4,aes(x=year,y=Estimate))+
  geom_line()+
  geom_ribbon(aes(ymin=`Q2.5`,ymax=`Q97.5`),fill="cadetblue",alpha=0.5,color=NA)+
  geom_point(mapping=aes(x=year,y=ilogit(logit_pALL)))+
  ylim(0,NA)
```

grid.arrange(p1,p2,p3,nrow=1)
